Modal Abstractions for Virtualizing Memory Addresses

### 2025-08-16 background.tex
check in some last-minute revision-tweaks that were never committed

Of particular interest for reasoning about virtual memory are modalities that permit \emph{naming} the alternate 
circumstances, prominently \emph{hybrid} modal logics~\cite{blackburn1995hybrid,areces2001hybrid}, which come equipped 
with assertions of the form $[\ell](P)$ indicating that $P$ is true in the specific alternate circumstance (Kripke world)
 named by the \emph{nominal} $\ell$. Note that a distinctive property of hybrid logics is that, rather than hiding
the points at which a modal assertion is evaluated inside the modality's definition, the choice of what world a modalized
assertion should be true in is chosen \emph{in the assertion itself}. This allows assertions to talk about not simply whether some other assertion
is true in some possible future or past world related in a fixed way to the current world, but to talk about \emph{arbitrary}
other worlds.
\add{This is somewhat different from the accessibility relation modalities more common in verification, but also well-established,
tracing its origin back to Arthur Prior~\cite{areces2006hybrid}, just as many temporal logics do.}

This explicit naming of alternate worlds increases the power of \add{propositional} modal logics~\cite{blackburn1995hybrid}, and is actually
necessary for completeness in classical separation logics~\cite{brotherston2014parametric}.
\add{However typically modal logics can always be ``compiled'' into logics with suitable quantifiers:
the \emph{standard translation}~\cite{blackburn2006modal} of propositional modal logics into first-order logic
has been thoroughly studied. \textsc{Iris}'s support for higher-order impredicative quantification means
this applies to our modalities as well --- they do not strictly speaking make \textsc{Iris} more expressive.
But while propositional modal logics are no more expressive than first-order logic (with just two variables!),
the primary goal of a modal logic is not raw expressive power but intuitive specifications and proof rules.
Temporal logics are used because they simplify specifications (vs. having explicit time variables
in every base asertion) and reasoning principles (because the modal correspondents of quantification over
time is used in a highly structured way).
Our logic offers a similar value proposition: natural specifications and reasoning principles which
are powerful enough for reasoning about virtual memory management, but without requiring pervasive tracking of
and explicit abstraction over address spaces.
}

>>>>

Of particular interest for reasoning about virtual memory are modalities that permit \emph{naming} the alternate 
circumstances, prominently \emph{hybrid} modal logics~\cite{blackburn1995hybrid,areces2001hybrid}, which come equipped 
with assertions of the form $[\ell](P)$ indicating that $P$ is true in the specific alternate circumstance (Kripke world)
 named by the \replace{\emph{nominal}}{term} $\ell$. Note that a distinctive property of hybrid logics is that, rather than hiding
the points at which a modal assertion is evaluated inside the modality's definition, the choice of what world a modalized
assertion should be true in is chosen \emph{in the assertion itself}. This allows assertions to talk about not simply whether some other assertion
is true in some possible future or past world related in a fixed way to the current world, but to talk about \emph{arbitrary}
other worlds.
\add{This is somewhat different from the accessibility relation modalities more common in verification, but also well-established,
tracing its origin back to Arthur Prior~\cite{areces2006hybrid}, just as many temporal logics do.}

This explicit naming of alternate worlds increases the power of \add{propositional} modal logics~\cite{blackburn1995hybrid}, and is \del{actually}
necessary for completeness in classical separation logics~\cite{brotherston2014parametric}.
\add{However typically modal logics can always be ``compiled'' into logics with suitable quantifiers:
the \emph{standard translation}~\cite{blackburn2006modal} of propositional modal logics into first-order logic
has been thoroughly studied. \textsc{Iris}'s support for higher-order impredicative quantification means
this applies to our modalities as well --- they do not strictly speaking make \textsc{Iris} more expressive.
But while propositional modal logics are no more expressive than first-order logic (with just two variables!),
the primary goal of a modal logic is not raw expressive power but intuitive specifications and proof rules.
Temporal logics are used because they simplify specifications (vs. having explicit time variables
in every base asertion) and reasoning principles (because the modal correspondents of quantification over
time is used in a highly structured way).
Our logic offers a similar value proposition: natural specifications and reasoning principles which
are powerful enough for reasoning about virtual memory management, but without requiring pervasive tracking of
and explicit abstraction over address spaces.
}

### 2025-08-16 intro.tex
check in some last-minute revision-tweaks that were never committed

\section{Introduction}
\label{sec:intro}
Virtual memory management lies at the core of modern OS kernel implementation. It is deeply intertwined with most other parts of a typical general-purpose OS kernel design, including scheduling, hardware drivers, and even the filesystem buffer cache. In writing the authoritative reference on the internals of the Solaris kernel, McDougall and Mauro went so far as to claim that ``\emph{the virtual memory sub-system can be considered the core of a Solaris instance, and the implementation of Solaris virtual memory affects just about every other subsystem in the operating system}''~\cite{mcdougall2006solaris}.
This makes rich support for verification the virtual memory management subsystem of an OS kernel critical to the correctness of every other piece of an OS or any software running atop it.

>>>>

\section{Introduction}
\label{sec:intro}
Virtual memory management lies at the core of modern OS kernel implementation. It is deeply intertwined with most other parts of a typical general-purpose OS kernel design, including scheduling, hardware drivers, and even the filesystem buffer cache. In writing the authoritative reference on the internals of the Solaris kernel, McDougall and Mauro went so far as to claim that ``\emph{the virtual memory sub-system can be considered the core of a Solaris instance, and the implementation of Solaris virtual memory affects just about every other subsystem in the operating system}''~\cite{mcdougall2006solaris}.
This makes \del{rich} support for verification the virtual memory management subsystem of an OS kernel critical to the correctness of every other piece of an OS or any software running atop it.
